---
title: "Additional Time Series Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional Time Series Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(tstoolbox)
library(ggplot2)
```

This vignette covers additional tools in **tstoolbox** beyond the core co-movement analysis functions:

- **Cross-correlation analysis** — Explore lagged correlations between series
- **Adstock transformations** — Apply and reverse advertising decay effects
- **Time series utilities** — Helpful functions for common tasks

## Cross-Correlation Analysis

Cross-correlation measures the correlation between two series at various lags. This helps identify:

- Whether series are related
- The optimal lag for the relationship
- Whether one series leads or follows another

### Creating a Cross-Correlation Table

```{r xcf-basic}
# Use European stock data
stocks <- as.data.frame(EuStockMarkets)

# Cross-correlation between DAX and CAC
xcf_result <- xcf(stocks, DAX, CAC)
head(xcf_result, 10)
```

The output shows:

- `lag`: The time offset (negative = DAX leads, positive = CAC leads)
- `x.corr`: The cross-correlation at that lag

### Visualising Cross-Correlation

Base R's `ccf()` produces functional but plain plots. `plot_xcf()` creates publication-ready ggplot2 visualisations:

```{r plot-xcf}
plot_xcf(stocks, DAX, CAC, title = "DAX vs CAC Cross-Correlation")
```

**Reading the plot**:

- Green bars: Positive correlation
- Red bars: Negative correlation
- The tallest bar indicates the lag with strongest relationship
- Bars near lag 0 suggest contemporaneous relationship

### Comparing Multiple Pairs

```{r xcf-multiple, fig.height=6}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# Quick base R comparisons
for (idx in c("SMI", "CAC", "FTSE")) {
  ccf(stocks$DAX, stocks[[idx]], main = paste("DAX vs", idx), lag.max = 20)
}
```

## Adstock Transformations

Adstock is a concept from marketing analytics that models the **carryover effect** of advertising. When you see an ad today, its effect doesn't disappear immediately — it decays over time.

### The Adstock Formula

$$\text{Adstock}_t = X_t + \lambda \cdot \text{Adstock}_{t-1}$$

Where:

- $X_t$ is the original value (e.g., ad spend)
- $\lambda$ is the decay rate (0 to 1)
- Higher $\lambda$ = longer-lasting effect

### Applying Adstock

```{r adstock-basic}
# Simulate weekly advertising spend
set.seed(42)
ad_spend <- c(100, 0, 0, 0, 200, 0, 0, 50, 0, 0, 0, 0)

# No decay (rate = 0): same as original
adstock(ad_spend, rate = 0)

# Light decay (rate = 0.3)
adstock(ad_spend, rate = 0.3)

# Heavy decay (rate = 0.7)
adstock(ad_spend, rate = 0.7)
```

### Visualising Adstock Effect

```{r adstock-visual}
# Compare different decay rates
rates <- c(0, 0.3, 0.5, 0.7)
results <- sapply(rates, function(r) adstock(ad_spend, rate = r))
colnames(results) <- paste0("rate_", rates)

# Plot
matplot(results, type = "l", lty = 1, lwd = 2,
        col = c("black", "blue", "orange", "red"),
        xlab = "Week", ylab = "Adstock Value",
        main = "Adstock Transformation at Different Decay Rates")
legend("topright", legend = paste("λ =", rates), 
       col = c("black", "blue", "orange", "red"), lty = 1, lwd = 2)
```

**Interpretation**:

- Black (rate=0): No carryover, values are unchanged
- Blue (rate=0.3): Light carryover, effects decay quickly
- Red (rate=0.7): Heavy carryover, effects persist for many periods

### Reversing Adstock

If you have adstocked data and need to recover the original values:

```{r reverse-adstock}
original <- c(100, 200, 300, 150, 200)

# Apply adstock
adstocked <- adstock(original, rate = 0.4)
adstocked

# Reverse to get original back
recovered <- reverse_adstock(adstocked, rate = 0.4)
recovered

# Verify
all.equal(original, recovered)
```

### Practical Example: Marketing Mix Modeling

```{r mmm-example}
# Simulate marketing data
set.seed(123)
n_weeks <- 52

marketing_data <- data.frame(
  week = 1:n_weeks,
  tv_spend = rpois(n_weeks, lambda = 50) * rbinom(n_weeks, 1, 0.3),
  digital_spend = rpois(n_weeks, lambda = 30),
  sales = 1000 + rnorm(n_weeks, sd = 50)
)

# Apply adstock with different rates for different channels
# TV has longer carryover than digital
marketing_data$tv_adstock <- adstock(marketing_data$tv_spend, rate = 0.6)
marketing_data$digital_adstock <- adstock(marketing_data$digital_spend, rate = 0.3)

head(marketing_data)

# Now these adstocked variables can be used in regression
# model <- lm(sales ~ tv_adstock + digital_adstock, data = marketing_data)
```

## Time Series Utilities

### Percentage Change

Calculate period-over-period percentage change:

```{r pc-change}
prices <- c(100, 110, 105, 115, 120)

# Default: 1-period change
pc_change(prices)

# 2-period change
pc_change(prices, lag = 2)
```

This is useful for:

- Calculating returns from prices
- Normalising series with different scales
- Growth rate analysis

### Returning k-th Date

Find specific dates in a vector:

```{r return-k-date}
library(lubridate)

dates <- ymd(c("2024-01-15", "2024-03-20", "2024-02-10", "2024-04-05"))

# Most recent date
return_k_date(dates, k = 1)

# 2nd most recent
return_k_date(dates, k = 2)

# 2nd oldest (decreasing = FALSE)
return_k_date(dates, k = 2, decreasing = FALSE)
```

### Time Series Aggregation

Aggregate time series data by time periods using dplyr operations:

```{r ts-summarise}
# Create daily data
set.seed(789)
daily_data <- data.frame(
  Date = seq(as.Date("2024-01-01"), as.Date("2024-03-31"), by = "day"),
  sales = rpois(91, lambda = 100),
  visitors = rpois(91, lambda = 500)
)

head(daily_data)

# Aggregate to monthly using base R
library(dplyr)
monthly <- daily_data %>%
  mutate(month = format(Date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(
    sales = sum(sales),
    visitors = sum(visitors)
  )
monthly

# Aggregate to weekly 
weekly <- daily_data %>%
  mutate(week = lubridate::floor_date(Date, "week")) %>%
  group_by(week) %>%
  summarise(
    sales = mean(sales),
    visitors = mean(visitors)
  )
head(weekly)
```

### Start-End Line

Generate a linear sequence between the first and last values:

```{r stend-line}
values <- c(100, 120, 90, 150, 200)

# Linear interpolation from start to end
stend_line(values)

# Useful for calculating deviation from trend
actual <- values
trend <- stend_line(values)
deviation <- actual - trend
data.frame(actual, trend, deviation)
```

### Sum of Lagged Differences (Fluctuation Score)

Measure how much a series fluctuates:

```{r sumlagdiff}
# Stable series
stable <- c(100, 101, 99, 100, 102, 98)
sumlagdiff(stable)

# Volatile series
volatile <- c(100, 150, 80, 200, 50, 180)
sumlagdiff(volatile)
```

Higher values indicate more volatility/fluctuation. This can be used to:

- Compare volatility across series
- Identify regime changes
- Flag unusual activity

## Combining Tools

### Example: Complete Marketing Analysis Pipeline

```{r complete-example, fig.height=6}
# Simulate weekly data
set.seed(456)
n <- 100

df <- data.frame(
  week = 1:n,
  ad_spend = rpois(n, 50) * rbinom(n, 1, 0.4),
  competitor_spend = rpois(n, 40) * rbinom(n, 1, 0.5)
)

# Apply adstock
df$ad_adstock <- adstock(df$ad_spend, rate = 0.5)
df$comp_adstock <- adstock(df$competitor_spend, rate = 0.5)

# Simulate sales influenced by both
df$sales <- 500 + 
  0.5 * df$ad_adstock - 
  0.3 * df$comp_adstock + 
  rnorm(n, sd = 30)

# 1. Basic co-movement between our ads and sales
cat("=== Ad Spend vs Sales Co-movement ===\n")
analyse_direction(df, ad_adstock, sales)

# 2. Cross-correlation
cat("\n")
plot_xcf(df, ad_adstock, sales, title = "Ad Spend vs Sales Cross-Correlation")
```

```{r complete-example-2}
# 3. Check for lead-lag
cat("\n=== Lead-Lag Analysis ===\n")
direction_leadlag(df$ad_adstock, df$sales, max_lag = 8)

# 4. Rolling relationship
plot_rolling_direction(
  df$ad_adstock, df$sales,
  window = 20,
  title = "Rolling Ad-Sales Relationship"
)
```

## Summary

| Category | Function | Purpose |
|----------|----------|---------|
| **Cross-Correlation** | `xcf()` | Create cross-correlation table |
| | `plot_xcf()` | Visualise cross-correlation |
| **Adstock** | `adstock()` | Apply decay transformation |
| | `reverse_adstock()` | Reverse decay transformation |
| **Utilities** | `pc_change()` | Percentage change |
| | `return_k_date()` | Get k-th date from vector |
| | `ts_summarise()` | Aggregate by time period |
| | `stend_line()` | Linear interpolation |
| | `sumlagdiff()` | Fluctuation score |

## See Also

- Main vignette: "Introduction to Co-movement Analysis" for the core analytical functions
- Function documentation: `?function_name` for detailed parameter descriptions
