---
title: "Introduction to Co-movement Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Co-movement Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(

collapse = TRUE,
comment = "#>",
fig.width = 7,
fig.height = 4
)
```

```{r setup}
library(tstoolbox)
library(ggplot2)
```
## What is Co-movement Analysis?

Co-movement analysis examines how two time series move together over time. Unlike correlation (which measures linear association), **directional co-movement** asks a simpler question:

> *"What proportion of the time do these series move in the same direction?"*

This is particularly useful when:

- You care about direction more than magnitude (e.g., "Did both stocks go up?")
- Series have different scales or volatilities
- You want an intuitive, interpretable measure
- You need to detect changing relationships over time

## The Dataset: European Stock Markets

We'll use the `EuStockMarkets` dataset that comes with R, containing daily closing prices of four European stock indices from 1991-1998:

- **DAX** (Germany)
- **SMI** (Switzerland)
- **CAC** (France)
- **FTSE** (UK)

```{r data-prep}
# Convert to data frame
stocks <- as.data.frame(EuStockMarkets)

# Create a date sequence (EuStockMarkets is daily data from 1991-1998)
# The time() function returns decimal years, so we convert appropriately
time_vals <- time(EuStockMarkets)
stocks$date <- as.Date("1991-01-01") + (time_vals - 1991) * 365.25

head(stocks)
```

## Basic Direction Analysis

### Understanding Direction

The `direction()` function converts a numeric series into directional changes:

```{r direction-basic}
# Simple example
prices <- c(100, 105, 103, 108, 108, 110)
direction(prices)
```

Each value shows whether the series went up ("Positive"), down ("Negative"), or stayed flat ("Equal") compared to the previous observation.

### Analysing Co-movement Between Two Series

Do German (DAX) and French (CAC) markets move together? Use `analyse_direction()`:
```{r analyse-direction}
analyse_direction(stocks, DAX, CAC)
```

The output tells us:

- The number of days where both indices moved in the same direction
- The proportion (as a percentage)
- This is returned as a tibble for further analysis

Let's compare all pairs:

```{r all-pairs}
# DAX vs all others
analyse_direction(stocks, DAX, SMI)
analyse_direction(stocks, DAX, FTSE)
```

**Interpretation**: European markets show strong co-movement (65-70%), which makes sense given economic integration.

## Statistical Testing

Is 70% co-movement statistically significant, or could it happen by chance?

### The Binomial Test

Under the null hypothesis of independence, two series have a 50% chance of moving in the same direction. We can test against this:

```{r direction-test}
result <- direction_test(stocks$DAX, stocks$CAC)
print(result)
```

The extremely small p-value confirms that DAX-CAC co-movement is highly significant.

### Alternative Testing Methods

For robustness, you can use permutation or bootstrap tests:

```{r alternative-tests, eval=FALSE}
# Permutation test (shuffles one series to create null distribution
direction_test(stocks$DAX, stocks$CAC, method = "permutation", n_sim = 1000)

# Bootstrap test (resamples to estimate uncertainty)
direction_test(stocks$DAX, stocks$CAC, method = "bootstrap", n_sim = 1000)
```

### One-Sided Tests

Test specific hypotheses:

```{r one-sided}
# Do they move together MORE than chance?
direction_test(stocks$DAX, stocks$CAC, alternative = "greater")
```

## Rolling Co-movement Analysis

Markets don't have constant relationships. The 2008 financial crisis famously showed that "correlations go to 1 in a crisis." Let's see how co-movement changes over time.

### Calculating Rolling Co-movement

```{r rolling-basic}
# 60-day rolling window
rolling_prop <- rolling_direction(stocks$DAX, stocks$CAC, window = 60)

# Add to data frame
stocks$dax_cac_rolling <- rolling_prop

# View summary
summary(rolling_prop)
```

### Visualising Rolling Co-movement

```{r plot-rolling}
plot_rolling_direction(
  stocks$DAX, 
  stocks$CAC, 
  window = 60,
  time = stocks$date,
  title = "DAX-CAC Rolling Co-movement"
)
```

**Reading the plot**:

- Values above 0.5 (green zone): Positive co-movement
- Values below 0.5 (red zone): Inverse movement
- The dashed line at 0.5 represents "chance" level

### Window Size Considerations

```{r window-comparison, fig.height=6}
# Compare different windows
windows <- c(20, 60, 120)

par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (w in windows) {
  roll <- rolling_direction(stocks$DAX, stocks$CAC, window = w)
  plot(stocks$date, roll, type = "l", 
 ylim = c(0.3, 1),
       main = paste("Window =", w, "days"),
       xlab = "Date", ylab = "Co-movement")
  abline(h = 0.5, lty = 2, col = "gray")
}
```

- **Short windows** (20): More volatile, captures short-term regime changes
- **Medium windows** (60): Good balance, ~3 months of trading
- **Long windows** (120): Smoother, shows persistent trends

## Lead-Lag Analysis

Does one market lead another? Perhaps US markets (not in this dataset) lead European markets, or maybe Germany leads France?

### Detecting Lead-Lag Relationships

```{r leadlag-basic}
# Does DAX lead or lag CAC?
ll_result <- direction_leadlag(stocks$DAX, stocks$CAC, max_lag = 10)
print(ll_result)
```

The `optimal_lag` tells us:

- **Negative lag**: First series (x) leads
- **Positive lag**: Second series (y) leads
- **Zero**: Contemporaneous (same-day) relationship

### Visualising Lead-Lag Structure

```{r plot-leadlag}
plot(ll_result)
```

The bar chart shows co-movement at each lag. The red dashed line marks the optimal lag.

### Example: Creating a Known Lead-Lag

Let's simulate data where x clearly leads y:

```{r simulated-leadlag}
set.seed(42)
n <- 200
x <- cumsum(rnorm(n))
y <- dplyr::lag(x, 3) + rnorm(n, sd = 0.5)  
# y follows x with 3-period lag

result <- direction_leadlag(x, y, max_lag = 6)
print(result)
plot(result)
```

The analysis correctly identifies that x leads y by 3 periods.

## Concordance Index

The Harding-Pagan concordance index is a formal measure used in economics to quantify business cycle synchronisation.

### Basic Usage

```{r concordance-basic}
conc <- concordance(stocks$DAX, stocks$CAC)
print(conc)
```

**Key outputs**:

- **Concordance**: Raw proportion of time in same state (like `analyse_direction`)
- **Expected**: What we'd expect under independence
- **Adjusted**: Transformed to [-1, 1] scale where 0 = independence

### Interpreting Adjusted Concordance

| Value | Interpretation |
|-------|----------------|
| 1 | Perfect positive co-movement |
| 0 | Independence (no relationship) |
| -1 | Perfect negative co-movement |

```{r concordance-examples}
# Strong positive co-movement
concordance(stocks$DAX, stocks$CAC)$adjusted

# What about inverse relationship?
concordance(stocks$DAX, -stocks$CAC)$adjusted
```

## Asymmetric Co-movement

Do markets co-move differently during bull vs bear markets? This is crucial for portfolio diversification — you want assets that diversify in *downturns*, not just overall.

### Testing for Asymmetry

```{r asymmetric-basic}
asym <- asymmetric_direction(stocks$DAX, stocks$CAC)
print(asym)
```

**Key outputs**:

- **upturn**: Co-movement when reference series is rising
- **downturn**: Co-movement when reference series is falling
- **asymmetry**: Difference (upturn - downturn)
- **p_value**: Is the asymmetry statistically significant?

### Choosing the Reference Series

The `reference` argument controls how upturns/downturns are defined:

```{r asymmetric-reference}
# Use DAX to define market regimes
asymmetric_direction(stocks$DAX, stocks$CAC, reference = "x")

# Use CAC to define market regimes
asymmetric_direction(stocks$DAX, stocks$CAC, reference = "y")
```

### Simulating Asymmetric Co-movement

```{r simulated-asymmetric}
set.seed(123)
n <- 500

# x is our reference series
x <- cumsum(rnorm(n))

# y follows x tightly in downturns, loosely in upturns
y <- numeric(n)
y[1] <- 0
for (i in 2:n) {
  if (x[i] < x[i-1]) {
    # Downturn: strong relationship
    y[i] <- y[i-1] + (x[i] - x[i-1]) * 0.9 + rnorm(1, sd = 0.1)
  } else {
    # Upturn: weak relationship
    y[i] <- y[i-1] + rnorm(1, sd = 1)
  }
}

asymmetric_direction(x, y)
```

This correctly identifies that co-movement is stronger during downturns.

## Complete Workflow Example

Let's put it all together with a full analysis:

```{r workflow, fig.height=8}
# 1. Basic co-movement
cat("=== Basic Co-movement ===\n")
analyse_direction(stocks, DAX, SMI)

# 2. Statistical significance
cat("\n=== Statistical Test ===\n")
test <- direction_test(stocks$DAX, stocks$SMI)
print(test)

# 3. Formal concordance measure
cat("\n=== Concordance Index ===\n")
conc <- concordance(stocks$DAX, stocks$SMI)
cat("Adjusted concordance:", round(conc$adjusted, 3), "\n")

# 4. Lead-lag structure
cat("\n=== Lead-Lag Analysis ===\n")
ll <- direction_leadlag(stocks$DAX, stocks$SMI, max_lag = 5)
print(ll)

# 5. Asymmetry
cat("\n=== Asymmetric Co-movement ===\n")
asym <- asymmetric_direction(stocks$DAX, stocks$SMI)
cat("Upturn co-movement:", round(asym$upturn * 100, 1), "%\n")
cat("Downturn co-movement:", round(asym$downturn * 100, 1), "%\n")

# 6. Rolling analysis (visual)
plot_rolling_direction(
  stocks$DAX, stocks$SMI, 
  window = 60, time = stocks$date,
  title = "DAX-SMI: Rolling 60-Day Co-movement"
)
```

## Summary

| Function | Purpose | Key Output |
|----------|---------|------------|
| `direction()` | Convert series to directional changes | Character vector |
| `analyse_direction()` | Basic co-movement proportion | Tibble with n, base, prop |
| `direction_test()` | Statistical significance | p-value, confidence interval |
| `rolling_direction()` | Time-varying co-movement | Numeric vector |
| `plot_rolling_direction()` | Visualise rolling co-movement | ggplot object |
| `direction_leadlag()` | Detect lead-lag relationships | Optimal lag, interpretation |
| `concordance()` | Formal concordance index | Raw and adjusted measures |
| `asymmetric_direction()` | Upturn vs downturn co-movement | Asymmetry measure, p-value |

## Next Steps

- See the "Additional Time Series Tools" vignette for cross-correlation, adstock transformations, and utility functions
- Check function documentation with `?function_name` for detailed parameter descriptions
